\color{navyblue} Differences-in-Differences with Spatial Spillover

### 2021-01-28 latex/paper/Spillover.tex
Edit introduction
Add Hannah's revisions

Empirical work in economics often considers treatment assigned by geographic boundaries such as cities, counties, and states. Potentially, the effect of these treatments do not stay within these boundaries whether it be from people crossing borders to access treatment (imperfect compliance) or from general equilibrium effects affecting neighboring areas. In the causal inference literature, the Stable Unit Treatment Value Assumption (SUTVA) is often assumed which says that a unit's potential outcome does not depend on the treatment status of any other unit \citep{Rubin_1980}. This article considers a common violation of SUTVA in a process I label `spatial spillovers' where `close' units' treatment assignment affects outcomes of nearby units.\footnote{Close can refer to many different things, e.g. geographic distance, economic distance such as supply chains, node distance in a graph, or social relationships in schools or cities.}

This paper identifies problems in estimation of treatment effects in the presence of spillovers and provides practical advice on using specifications that are robust to spillovers. To do this, I first introduces a potential outcome framework that explicitly models spatial spillovers following \citet{Vazquez-Bare_2019}. Typically researchers are interested in the `direct' effect on the treated units (sometimes called the partial equilibrium effect or the `local' effect). In the presence of spatial spillovers, difference-in-differences estimates identify the direct effect of treatment (the estimand of interest) plus two additional bias terms resulting from the spillovers.

The intuition for these two sources is as follows. First, untreated units that are `close' to treated units experience effects of treatment and therefore these `control' units fail to identify the counterfactual trend. When estimating by difference-in-differences, the spillover onto the `close' control units is averaged into the untreated change in outcomes. In this case, the spillover is subtracted from the estimated treatment effect and biases the estimate in the opposite sign of the spillover effect. For example, consider a factory opening in a county. Agglomeration economies suggest that neighboring counties would also benefit from knowledge spillovers and improved access to supply chain \citep{Duranton_Puga_2003}. The treatment effect estimate is negatively biased because the change in outcome in neighboring counties is higher than it would be absent treatment. Researchers would therefore underestimate the positive benefits of treatment.\footnote{If the policy maker had a welfare function that included both treated and neighboring counties, they would fail to count the positive benefits of the control units and they would underestimate the positive benefits on the treated units. Difference-in-differences estimation would result in double undercounting.}

Second, changes in treated units' outcomes reflect the effect of their own treatment status and the effect from the treatment status of ``close'' units. The spillover is added to the treated units' change in outcomes. Therefore the bias of treatment effect is the same sign as the sign of the spillover. Continuing with our example, two factory openings in neighboring counties might cause the benefit of each individual factory to increase due to agglomeration forces. The estimated treatment effect will count the `direct` effect of the factory opening as well as the benefit of the agglomeration economies. Therefore the treatment effect will be positively biased due to the positive spillover onto also treated units.

These biases are explicit functions of potential outcomes and can be used by researchers to sign or bound the bias using insight from their economic context. The magnitude of the bias depends on two factors: the size of the spillovers and the number of units affected by spillovers. If spillovers are quite large in magnitude or spread far over distance, then the bias will generally be large. This result of this paper enables researchers to give bounds for it under assumptions about the size of the spillover effect and on number of units affected. I use Monte Carlo simulations to quantify the magnitude of bias in various spillover settings.

Despite the problem of spatial spillovers being common across many settings, \citet{Berg_Streitz_2019} document that little empirical analysis takes the problem seriously. In a survey of eight top economics and finance journals in 2017, they find that only 21 articles out of 108 that run differencse-in-differences estimation discuss spillovers.\footnote{The articles surveyed are American Economic Review, Econometrica, the Journal of Political Economy, the Quarterly Journal of Economics, the Review of Economic Studies, the Journal of Finance, the Journal of Financial Economics, and the Review of Financial Studies.} Of those 21, only eight include spillovers in their empirical specification to prevent bias in the estimation of treatment effects. The other papers usually will first estimate the direct effect without controlling for spillovers and then later on seperately estimate the spillover effects. This paper shows that not controlling for spillovers produces biased estimates of the direct effect.

Of the eight papers that include spillovers directly in their estimation strategy, six simply drop the control units they suspect to be affected by spillovers. Since this is a common strategy by researchers, I use Monte Carlo simulations to show that this method is effective in the case where there is \textit{only} spillovers onto control units. In the presence of spillover onto other treated units, the estimates will still be biased for the direct effect. Dropping control unit observations, however, makes estimates less precise, so parameterizing the spillover function may be a better strategy especially in the case where a high proportion of control units would need to be removed.

Removing both sources of bias requires a researcher to take a stance and parameterize a functional form for the spillover function. The particular functional form depends on the context of the application. However, there are a few practical considerations. First, the researcher should argue whether spillovers on control units and on treated units exist. Second, they should argue what the maximum distance these spillovers can occur. Last, they should consider if spillovers are additive in the number of nearby units that are treated.

% @TODO: Discuss effectiveness under misspecification of spillovers.

@@

The literature considering spillovers primarily has focused on ``within-group'' spillovers where units are in distinct groups and outcomes depend on the treatment status within the observation's group only.\footnote{For example, researchers consider peer effects within schools. In this case, each school constitutes a group and it is assumed that outcomes do not depend on other schools' treatment status, i.e. SUTVA is satisfied between-groups. Estimation of spillover effects in this case is done in a `partial intereference' framework that compares ``control'' units in a partially treated group with control units in completely untreated groups. \citet{Angelucci_DiMaro_2016} provides a summary for estimation of spillover effects in within-group settings. Examples in the literature include: \citet{Halloran_Struchiner_1995} considers community-vaccine rates in epidemology; \citet{Miguel_Kremer_2004} consider deworming programs in Kenyan schools; \citet{Sobel_2006} considers interference in the Moving to Opportunity Program; and \citet{Angrist_2014} studies the context of school peer effects.} 
However, spatial settings such as counties do not feature distinct groups which allow for the researcher to seperate areas into distinct non-overlapping groups (a county is connected to all other counties by some chain of neighbors). My paper considers a more general form of interference. Identification in this framework requires the researcher to either identify control units that are free from spillovers (far-away from treatment) or to parameterize how control units are affected by treatment.

@@

\citet{Sävje_Aronow_Hudgens_2019} consider a similar potential outcome framework as \citet{Vazquez-Bare_2019}, but they instead include spillovers as a part of their definition of ``treatment effect''. That is, a treated unit's treatment effect is defined as the direct effect and the general equilibrium spillover effects. Then, the authors describe how the ATE estimated by difference-in-differences averages over both individual heterogeneity and, central to their contribution, over heterogeneity in the exposure mapping, $h(\vec{D}, i)$. My paper develops a strategy to allow researchers to seperately identify treatment effects and spillovers.\footnote{These estimates can be combined afterwards to estimate their definition of average treatment effect. The further advantage is that `net' treatment effects can be estimated at different levels of exposure which is more relevant for future policy decisions.}

@@

My paper derives an explicit form for this bias in terms of general potential outcomes which allows my results to capture many different forms of spillovers. For the above papers, if I assume the particular functional forms for potential outcomes, I arrive at the same bias equation as theirs if they have one derived explicitly.

@@

The rest of the paper is structured as follows. Section \ref{sec:po_framework} presents the potential outcomes framework, defines the estimand of interest, and shows the resulting bias from estimating a classical difference-in-differences model. Section \ref{sec:monte_carlo} presents Monte Carlo simulations to illlustrate the bias result and to evaluate currently used solutions in the literature. Section \ref{sec:tva} presents an application for evaluating the Tennessee Valley Authority program.

>>>>

Empirical work in economics often considers treatment assigned by geographic boundaries such as cities, counties, and states. Potentially, the effect of these treatments do not stay within these boundaries whether it be from people crossing borders to access treatment (imperfect compliance) or from general equilibrium effects influencing neighboring areas. Despite the problem of spatial spillovers being common across many settings, \citet{Berg_Streitz_2019} document that little empirical analysis takes the problem seriously. In a survey of eight top economics and finance journals in 2017, they find that only 21 articles out of 108 that run difference-in-differences estimation discuss spillovers.\footnote{The articles surveyed are American Economic Review, Econometrica, the Journal of Political Economy, the Quarterly Journal of Economics, the Review of Economic Studies, the Journal of Finance, the Journal of Financial Economics, and the Review of Financial Studies.} In this paper, I show that failing to account for spatial spillovers can lead to poorly identified estimates of treatment effects.

In the causal inference literature, researchers assume that a unit's potential outcome does not depend on the treatment status of any other unit -- what \citet{Rubin_1980} calls the Stable Unit Treatment Value Assumption (SUTVA). This article considers a common violation of SUTVA in a process I label `spatial spillovers' where the treatment assignment of `close' units' affects outcomes of nearby units.\footnote{Close can refer to many different things, e.g. geographic distance, economic distance such as supply chains, node distance in a graph, or social relationships in schools or cities.} Typically researchers are interested in the `direct effect' on the treated units.\footnote{This is sometimes called the `partial equilibrium' effect or the `local effect'.} In the presence of spatial spillovers, I show using a potential outcomes framework that the difference-in-differences estimate identifies the direct effect of treatment (the estimand of interest) plus two additional bias terms resulting from the spillovers.

The intuition for the two bias terms resulting from spillovers is as follows. First, untreated units that are `close' to treated units experience effects of treatment and therefore these `control' units fail to identify the counterfactual trend. When estimating by difference-in-differences, the spillover onto the `close' control units is averaged into the untreated units' change in outcomes. In this case, the spillover is subtracted from the estimated treatment effect and biases the estimate in the opposite sign of the spillover effect. For example, consider a factory opening in a county. Agglomeration economies suggest that neighboring counties would also benefit from knowledge spillovers and improved access to supply chain \citep{Duranton_Puga_2003}. The treatment effect estimate is negatively biased because the change in outcome in neighboring counties is higher than it would be absent treatment. Researchers would therefore underestimate the positive benefits of treatment.\footnote{If the policy maker had a welfare function that included both treated and neighboring counties, they would fail to count the positive benefits of the control units and they would underestimate the positive benefits on the treated units. Difference-in-differences estimation would therefore result in double undercounting.}

Second, changes in treated units' outcomes reflect the effect of their own treatment status and the effect from the treatment status of ``close'' units. The spillover onto other treated units is averaged into the treated units' change in outcomes. Therefore the bias of treatment effect is the same sign as the sign of the spillover. Continuing with our example, two factory openings in neighboring counties might cause the benefit of each individual factory to increase due to agglomeration forces. The estimated treatment effect will count the `direct` effect of the factory opening as well as the benefit of the agglomeration economies. Therefore the treatment effect will be positively biased due to the positive spillover onto also treated units.

These biases can be written as functions of the potential outcomes in my framework and can be used by researchers to sign or bound the bias using insight from their economic context. The magnitude of the bias depends on two factors: the size of the spillovers and the number of units affected by spillovers. If spillovers are quite large in magnitude or spread far over distance, then the bias will generally be large. Th results of this paper enable researchers to estimate bounds for it by making assumptions about the average size of the spillover effect and the number of units affected. I use Monte Carlo simulations to quantify the magnitude of bias in various spillover settings.

In the survey by \citet{Berg_Streitz_2019}, only eight of the 21 papers that discuss spillovers control for this potential source of bias in their primary empirical specification to prevent bias in the estimation of treatment effects. The other thirteen papers usually first estimate the direct effect without controlling for spillovers and then seperately estimate the spillover effects. This paper shows that not controlling for spillovers produces biased estimates of the direct effect.

Of the eight papers that include spillovers directly in their estimation strategy, six simply drop the control units they suspect to be affected by spillovers. Since this is a common strategy used by researchers, I use Monte Carlo simulations to show that this method is effective in the case where spillovers \textit{only} occur onto control units. In the presence of spillover onto also treated units, the estimates for the direct effect will still be biased. Dropping control unit observations, however, makes estimates less precise, so parameterizing the spillover function may be a better strategy especially in the case where a high proportion of control units would need to be removed.

Removing both sources of bias requires a researcher to take a stance and parameterize a functional form for the spillovers. The particular functional form depends on the context of the application. However, there are a few practical considerations. First, the researcher should argue whether spillovers on control units and on treated units exist. Second, they should argue what the maximum distance these spillovers can occur. Last, they should consider if spillovers are additive in the number of nearby units that are treated.

In section \ref{sec:monte_carlo}, I use Monte-Carlo simulations to emphasize when misspecificied spillovers still remove most of the bias and when they fail to do so.

% @TODO: Discuss effectiveness under misspecification of spillovers (correct distance and additive/non-additive).

@@

The literature considering spillovers primarily has focused on ``within-group'' spillovers where units are in distinct groups and outcomes depend on the treatment status within the observation's group only.\footnote{For example, researchers consider peer effects within schools. In this case, each school constitutes a group and it is assumed that outcomes do not depend on other schools' treatment status, i.e. SUTVA is satisfied between-groups. Estimation of spillover effects in this case is done in a `partial intereference' framework that compares ``control'' units in a partially treated group with control units in completely untreated groups. \citet{Angelucci_DiMaro_2016} provides a summary for estimation of spillover effects in within-group settings. Examples in the literature include: \citet{Halloran_Struchiner_1995} consider community-vaccine rates in epidemology; \citet{Miguel_Kremer_2004} consider deworming programs in Kenyan schools; \citet{Sobel_2006} considers interference in the Moving to Opportunity Program; and \citet{Angrist_2014} studies the context of school peer effects.} However, spatial settings such as counties do not feature distinct groups which allow for the researcher to separate areas into distinct non-overlapping groups (a county is connected to all other counties by some chain of neighbors). My paper considers a more general form of interference. Identification in this framework requires the researcher to either identify control units that are free from spillovers (far-away from treatment) or to parameterize how control units are affected by treatment.

@@

\citet{Sävje_Aronow_Hudgens_2019} consider a similar potential outcome framework as \citet{Vazquez-Bare_2019}, but they instead include spillovers as a part of their definition of ``treatment effect''. That is, a treated unit's treatment effect is defined as the direct effect \textit{and} the general equilibrium spillover effects. Then, the authors describe how the average treatment effect (ATE) estimated by difference-in-differences averages over both individual heterogeneity and, central to their contribution, over heterogeneity in the exposure mapping, $h(\vec{D}, i)$. My paper develops a strategy to allow researchers to seperately identify treatment effects and spillovers.\footnote{These estimates can be combined afterwards to estimate their definition of average treatment effect. The further advantage is that `net' treatment effects can be estimated at different levels of exposure which is more relevant for future policy decisions.}

@@

My paper derives an explicit form for this bias in terms of general potential outcomes which allows my results to capture many different forms of spillovers. For the above papers, if I assume the particular functional forms for potential outcomes, I arrive at the same bias equation as theirs if they have one derived explicitly. My paper also advances the literature by considering estimation of direct effects and spillover effects in event-study framework which allows for the common occurence of staggered treatment adoption.

@@

The rest of the paper is structured as follows. Section \ref{sec:po_framework} presents the potential outcomes framework, defines the estimand of interest, and shows the resulting bias from estimating a classical difference-in-differences model. Section \ref{sec:monte_carlo} presents Monte Carlo simulations to illlustrate the bias result and to evaluate currently used solutions in the literature. Section \ref{sec:tva} presents an application for evaluating the Tennessee Valley Authority program. Section \ref{sec:event_study} discusses briefly how to incorporate spillovers into event-study estimation following the insights from \citet{Callaway_Sant’Anna_2020}.
